---
title: "Final STAT 341 Report"
authors: Celia Carey, Noelle Haviland, Heewon Yi
format: html 
editor: visual 
embed-resources: true 
---

```{r}
#| include: false 
library(dagitty) 
library(CalvinBayes)
library(tidybayes)
library(tidyverse)
library(dplyr)
library(rstan)
library(bayesplot)
```

## Read in dataset

```{r}
library(readxl) 
exchange <- read_excel("fredgraph.xlsx")  
glimpse(exchange)
```

Variable Standardization

```{r}
exchange <- exchange |>   
  rename(FedFundsRate = "DFF (Federal Funds rate)",          
         RecessionIndicator = "BRARECDM (Brazilian recession indicator)",          
         JobPostings = "IHLIDXUS (Job Posting)",          
         Reals_USD_Exchange = "DEXBZUS (exchange rate)") |>   
  mutate(FedFundsRate_scaled = as.numeric(scale(FedFundsRate)),          
         JobPostings_scaled = as.numeric(scale(JobPostings)),          
         RecessionIndicator = as.numeric(if_else(RecessionIndicator == "0", 1, 2)))

exchange <- exchange |>
  drop_na(Reals_USD_Exchange)
```

```{r}
exchange <- exchange |>
  select(FedFundsRate_scaled, JobPostings_scaled, Reals_USD_Exchange, RecessionIndicator)
```

```{r}
# Compose data for stan model
stan_exchange <- compose_data(exchange)
```

## Research Question

To what degree does the US's Fed have on the Brazilian Reals to USD EXR based on their control over the Federal Funds Rate?

## Where is the data from?

The data is from the FRED website and we're looking at how the Brazilian Reals to USD Spot Exchange Rate (https://fred.stlouisfed.org/series/DEXBZUS#, Data Updated: 2025-04-07) is affected by the:

-   US Federal Funds Effective Rate, Percent, Daily, Not Seasonally Adjusted (https://fred.stlouisfed.org/series/DFF#, Data Updated: 2025-04-07)

-   OECD based Recession Indicators for Brazil from the Peak through the Trough, +1 or 0, Daily, Not Seasonally Adjusted (https://fred.stlouisfed.org/series/BRARECDM#, Data Updated: 2022-11-10)

-   Job Postings on Indeed in the United States, Index Feb, 1 2020=100, Daily, Seasonally Adjusted (https://fred.stlouisfed.org/series/IHLIDXUS#, Data Updated: 2025-04-08)

We first acquired data on April 3, but ultimately changed our minds and downloaded data on April 8 that we are planning to use. The data comes from the FRED website and is identified by an alphacode title (DEXBZUS, DFF,BRARECDM, and IHLIDXUS).

To format it in excel after downloading, we lined up the daily data that came in two sheets and placed the columns next to one another.

## Causal Diagram

```{r}
causal_diagram <- dagitty("dag{   
                          FedFundsRate -> Reals_USD_Exchange;  
                          RecessionIndicator -> Reals_USD_Exchange;   
                          JobPostings -> Reals_USD_Exchange }")  
gg_dag(causal_diagram,        
       size = 60,        
       highlight = 'Reals_USD_Exchange')
```

## Causal Diagram Explanation

-   The Federal Funds Effective Rate, Percent, Daily, Not Seasonally Adjusted is a measure of the daily average interest rate that banks lend the money they have in reserve to one another after hours as a percent.

-   The OECD based Recession Indicators for Brazil from the Peak through the Trough, +1 or 0, Daily, Not Seasonally Adjusted has two values: 1 and 0. A 1 indicates that Brazil is experiencing a period of expansion and the 0 indicates a period of recession.

-   The last factor: Job Postings on Indeed in the United States, Index Feb, 1 2020=100, Daily, Seasonally Adjusted represents seasonally adjusted job postings with an average over the last 7 days.

### Model Description

$$ \text{Reals USD Exchange} \sim \text{Gamma distribution}(\alpha_{[i]}, \lambda) $$

$$ \alpha_{[i]} = \beta_0+\beta_1*\text{FedFundsRate}+\beta_{\text{RecessionIndicator[i]}}+\beta_3*\text{JobPostings} $$

$$ \beta_0\sim\text{Normal}(.76028,.65) $$

$$ \beta_1\sim\text{Normal}(1.5, 2) $$

$$ \beta_2\sim\text{Normal(.5, .5)} $$

$$ \beta_3\sim\text{Normal(125, 30)} $$

$$ \lambda\sim\text{exp}(0,3) $$

#### Rationale for Priors

Response Variable: Reals to USD Exchange Rate

Predictor Variable: US Federal Funds rate ($\beta_1$) https://www.grumpy-economist.com/p/do-higher-interest-rates-raise-the An increase to the Federal Funds Rate should create an initial boost in the value of the dollar. With a higher value assigned to the USD, the exchange rate should increase. Therefore, this should be a positive relationship.

Predictor Variable: US Job Postings According to the Federal Reserve Bank of New York, ($\beta_3$) "rise in employment strengthens the foreign exchange value of the dollar". Therefore, also should be a positive relationship.

Categorical Variable: Brazil Recession Indicator ($\beta_{\text{RecessionIndicator[i]}}$) During a recession, the money that used to be able to purchase, say two gallons of milk, will now only be able to purchase a quarter gallon. This indicates that the currency has become less valueable and should therefore even out in relation to another currency, hence a negative relationship to the exchange rate.

## Prior Predictive Distribution

```{r}
n_sim <- 100  
prior_pred_dists <- tibble(  
  b0 = rnorm(n_sim, mean = 0.76028, sd = 0.65),
  b1 = rnorm(n_sim, mean = 1.5, sd = 2),   
  b2 = rnorm(n_sim, mean = 0.5, sd = 0.5),   
  b3 = rnorm(n_sim, mean = 125, sd = 30),     
  sigma = rexp(n_sim, rate = 3),   
  sim_number = c(1:n_sim)   )  

glimpse(prior_pred_dists)
```

```{r}
# Create prior predictive draws
prior_pred_dists <- prior_pred_dists |>
  mutate(
    mu = b0 + b1 + b2 + b3,
    y_sim = rgamma(n_sim, shape = mu^2 / sigma^2, rate = mu / sigma^2)
  )

glimpse(prior_pred_dists)

gf_dens(~ y_sim, data = prior_pred_dists, color = "darkblue") |>
  gf_labs(
    title = "Prior Predictive Density of Simulated Outcomes",
    x = "Simulated y",
    y = "Density"
  ) + 
  theme_bw()
```

## Model Fitting

```{r}
#| eval: false
exchange_fitted_model <- stan(file = 'finalproject.stan',
                              data = stan_exchange,
                              chains = 4,
                              iter = 3000,
                              warmup = 300,
                              refresh = 0)
```

```{r}
#| eval: false
saveRDS(exchange_fitted_model, file="exchange_fitted_model.RDS")
```

```{r}
exchange_fitted_model <- readRDS("exchange_fitted_model.RDS")
```

## Chain Diagnostics

```{r}
mcmc_trace(exchange_fitted_model,
           pars = vars(starts_with("b"), "sigma"))
```

The chains of this trace plot are stationary and well-mixing, except in the plot for lp\_\_. They stay within a fairly narrow range and overlap, but do not match exactly.

```{r}
mcmc_rank_overlay(exchange_fitted_model)
```

The chains of this trace rank plot are also stationary (they all mostly stay in the range of 100-150). The chains overlap, but are not directly on top of each other.

```{r}
exchange_fitted_model
```

Our Rhat value is exactly 1, indicating that the model has converged and we used an appropriate number of post-warmup iterations. Our n_eff values are quite high, indicating that we have plenty of effective samples because we used a sufficient amount of warmup iterations.

## Posterior Predictive Distribution

```{r}
psamp <- as.data.frame(exchange_fitted_model)
glimpse(psamp)

gf_dens(~b0, data = psamp)
```

```{r}
gf_dens(~b1, data = psamp)
```

```{r}
psamp <- psamp |>   
  rename(b2_1 = "b2[1]",          
         b2_2 = "b2[2]")
gf_dens(~b2_1, data = psamp)
```

```{r}
gf_dens(~b2_2, data = psamp)
```

```{r}
gf_dens(~b3, data = psamp)
```

```{r}
gf_dens(~sigma, data = psamp)
```

## Model Comparison

```{r}
rethinking::PSIS(exchange_fitted_model)
```

## Conclusions

The Federal Funds Rate had the most impact on the exchange rate from Reals to U.S. dollars. We originally predicted that the job postings would have a positive effect on the exchange rate, but our model shows us a negative effect which we have not found a reason for. The recession indicator had a positive impact, but it is drastically less than the impact of the federal funds rate.

Another factor to consider would be that there are two ways that an EXR can be changed. From each currency's end, the impact on the EXR can be cancelled out or drastically magnified when there are changes to the economic conditions of both countries. Apart from the data that we gathered, we did not additionally consider single events such as political strife that may have created an anomaly in the EXR.

## Citations

Cochrane, J. H. (2024, September 16). Do higher interest rates raise the exchange rate? The Grumpy Economist. https://www.grumpy-economist.com/p/do-higher-interest-rates-raise-the

Harris, E., & Zabka, N. M. (1995). The employment report and the dollar. Number, 1. https://doi.org/10.2139/ssrn.1001251
